---
title: "Data Analysis"
author: "Danielle Herrington"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: lumen
---

```{r setup, include=FALSE}
# Set options for code chunks
knitr::opts_chunk$set(
  echo = TRUE,          
  message = FALSE,      
  warning = FALSE,      
  fig.width = 6,        
  fig.height = 4        
)

# Set a CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com"))

# Install required packages (if not already installed)
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("janitor", quietly = TRUE)) install.packages("janitor")
if (!requireNamespace("lubridate", quietly = TRUE)) install.packages("lubridate")
if (!requireNamespace("scales", quietly = TRUE)) install.packages("scales")
if (!requireNamespace("leaflet", quietly = TRUE)) install.packages("leaflet")


# Load libraries
library(tidyverse)
library(janitor)
library(lubridate)
library(scales)
library(leaflet)

# Load the data
bike_rides_cleaned <- read_csv("/Users/danielleherrington/Documents/Cyclistic/Cyclistic Case Study 2/cyclistic-project/data/bike_rides_cleaned.csv")
```

# Visualising the Number of Rides by Weekday and Customer Type
```{r rides by weekday, echo=TRUE, message=FALSE, warning=FALSE}
bike_rides_cleaned %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%  # Extract the weekday from 'started_at'
  group_by(member_casual, weekday) %>%
  summarise(
    number_of_rides = n(),  # Count the total number of rides
    average_duration = mean(ride_length, na.rm = TRUE)  # Calculate the average ride duration
  ) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(
    x = weekday,  # Weekday on the x-axis
    y = number_of_rides,  # Total number of rides on the y-axis
    fill = member_casual  # Customer type represented by color
  )) +
  geom_col(position = "dodge") +  # Create a grouped bar chart
  geom_text(
    aes(label = scales::comma(number_of_rides)),  # Annotate bars with ride counts
    position = position_dodge(width = 0.9),  # Position the text above the bars
    vjust = -0.5,  # Adjust text vertically
    size = 3  # Set label size
  ) +
  labs(
    title = "Total Number of Rides by Weekday and Customer Type",  # Title of the plot
    x = "Day of the Week",  # Label for x-axis
    y = "Total Number of Rides",  # Label for y-axis
    fill = "Customer Type"  # Legend title for the fill color
  ) +
  scale_y_continuous(labels = scales::comma) +  # Format y-axis labels with commas
  theme_minimal() +  # Apply a minimal theme for clean aesthetics
  coord_flip()  # Flip coordinates for better readability
```

# Visualising the Number of Rides by Month and Customer Type
```{r rides by month}
bike_rides_cleaned %>%
  mutate(month = month(started_at, label = TRUE, abbr = TRUE)) %>%
  group_by(month, member_casual) %>%
  summarise(total_trips = n(), .groups = 'drop') %>%
  ggplot(aes(x = total_trips, y = month, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = scales::comma(total_trips)),
            position = position_dodge(width = 0.9),
            hjust = -0.1, size = 3) +
  labs(
    title = "Total Number of Rides by Month and Customer Type",
    x = "Total Number of Rides",
    y = "Month",
    fill = "Customer Type"
  ) +
  theme_minimal() +
  scale_x_continuous(labels = scales::comma)
```

# Calculating the Average Ride Length by Customer Type
```{r avg ride length}
avg_ride_length <- bike_rides_cleaned %>%
  group_by(member_casual) %>%
  summarise(average_duration = mean(ride_length, na.rm = TRUE)) 
```

# Visualising the Average Ride Length by Customer Type
```{r visualising avg ride length}
ggplot(avg_ride_length, aes(x = member_casual, y = average_duration, fill = member_casual)) +
  geom_bar(stat = "identity", show.legend = FALSE) +  # Use 'identity' to plot the summarised values
  labs(
    title = "Average Ride Length by Customer Type",
    x = "Customer Type",
    y = "Average Ride Duration (seconds)"
  ) +
  theme_minimal() +  # Clean theme for the plot
  scale_y_continuous(labels = scales::comma)  # Format y-axis labels with commas for readability
```


# Hourly Ride Analysis
```{r hourly ride analysis}

# Step 1: Extract the hour from 'started_at' and create a new column for hour
bike_rides_cleaned <- bike_rides_cleaned %>%
  mutate(hour_of_day = hour(started_at))  # Extract hour of the day

# Step 2: Group by hour of day and member type
ride_summary_by_hour <- bike_rides_cleaned %>%
  group_by(hour_of_day, member_casual) %>%
  summarise(number_of_rides = n())  # Count of rides for each hour

# Step 3: Plot the trip frequency by hour as a line graph
ggplot(ride_summary_by_hour, aes(x = hour_of_day, y = number_of_rides, color = member_casual, group = member_casual)) +
  geom_line(linewidth = 1) +  # Line graph
  geom_point(size = 2) +  # Points on the lines for visibility
  labs(
    title = "Trip Frequency by Hour for Customer Types",
    subtitle = "Number of trips by hour for casual members vs members",
    x = "Hour of the Day",
    y = "Number of Rides",
    color = "Customer Type"
  ) +
  scale_x_continuous(breaks = 0:23) +  # Ensure every hour is represented on the x-axis
  scale_y_continuous(labels = label_number(accuracy = 1)) + 
  theme_minimal()
```

# Analysing Ride Frequency by Day and Hour with a Heatmap
```{r ride heatmap analysis}

# Step 1: Extract the day of the week and hour of the day
bike_rides_cleaned <- bike_rides_cleaned %>%
  mutate(
    day_of_week = wday(started_at, label = TRUE, abbr = TRUE),  # Day of week (abbreviated names)
    hour_of_day = hour(started_at)  # Hour of the day
  )

# Step 2: Group by member type, day of week, and hour of day
ride_summary_heatmap <- bike_rides_cleaned %>%
  group_by(member_casual, day_of_week, hour_of_day) %>%
  summarise(ride_count = n()) %>%  # Count rides in each group
  ungroup()

# Step 3: Create the heatmap
ggplot(ride_summary_heatmap, aes(x = hour_of_day, y = day_of_week, fill = ride_count)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C", name = "Ride Count") +  # Color scale for the ride frequency
  facet_wrap(~ member_casual, ncol = 1) +  # Separate heatmaps for each rider type
  labs(
    title = "Ride Frequency by Day of the Week and Hour of the Day",
    subtitle = "Comparing usage patterns for Casual Riders and Members",
    x = "Hour of the Day",
    y = "Day of the Week"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),  # Increase facet label size for clarity
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )
```



# Visualising Top 5 Starting Stations (Per Customer Type) 
```{r top 5 starting stations}
# Group by customer type and starting station, count trips, filter out empty stations
top_starting_stations <- bike_rides_cleaned %>%
  filter(!is.na(start_station_name) & start_station_name != "") %>%  # Filter out NAs and empty strings
  group_by(member_casual, start_station_name) %>%  # Group by member type and start station
  summarise(trip_count = n(), .groups = "drop") %>%  # Count the number of trips
  arrange(desc(trip_count)) %>%  # Sort by trip count in descending order
  group_by(member_casual) %>%  # Group by member type to ensure top 10 per type
  top_n(5, trip_count) %>%  # Select top 5 starting stations per customer type
  ungroup()  # Remove grouping for further processing


# Plot the data with labels inside the bars
ggplot(top_starting_stations, aes(x = reorder(start_station_name, trip_count), y = trip_count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +  # Side-by-side bars for each member type
  coord_flip() +  # Flip the chart for horizontal bars
  labs(title = "Top 5 Starting Stations by Customer Type",
       x = "Starting Station",
       y = "Number of Rides",
       fill = "Customer Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis text
  facet_wrap(~ member_casual) +  # Facet by customer type
  geom_text(aes(label = trip_count), 
            position = position_dodge(width = 0.8),  # Align text with bars
            hjust = 1,  # Horizontally center the text inside the bars
            size = 4)  # Adjust the text size

```

# Visualising Top 5 Starting Stations (Per Customer Type) on a Leaflet Map
```{r top 5 starting stations leaflet map}
install.packages("leaflet")
library(leaflet)

# Step 1: Calculate top starting stations without including coordinates in the grouping
top_starting_stations <- bike_rides_cleaned %>%
  filter(!is.na(start_station_name) & start_station_name != "") %>%
  group_by(start_station_name, member_casual) %>%
  summarise(number_of_rides = n(), .groups = "drop") %>%
  arrange(desc(number_of_rides)) %>%
  group_by(member_casual) %>%
  slice_max(number_of_rides, n = 5)  # Select top 5 stations per customer type

# Step 2: Merge coordinates with the top 5 stations using distinct coordinates for each station
top_starting_stations_coords <- top_starting_stations %>%
  left_join(
    bike_rides_cleaned %>%
      select(start_station_name, start_lat, start_lng) %>%
      distinct(start_station_name, .keep_all = TRUE), # Ensure unique station coordinates
    by = "start_station_name"
  )

# Step 3: Create the Leaflet map with coordinates now reattached
leaflet(data = top_starting_stations_coords) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~start_lng, lat = ~start_lat,
    radius = ~sqrt(number_of_rides) / 8,  # Adjust radius for visualisation
    color = ~ifelse(member_casual == "member", "blue", "green"),  # Color by customer type
    fill = TRUE, fillOpacity = 0.5,
    popup = ~paste("Station:", start_station_name,
                   "<br>Customer Type:", member_casual,
                   "<br>Number of Rides:", number_of_rides)
  ) %>%
  addLegend(position = "bottomright", colors = c("blue", "green"),
            labels = c("Member", "Casual"), title = "Customer Type") %>%
  setView(lng = mean(top_starting_stations_coords$start_lng, na.rm = TRUE), 
          lat = mean(top_starting_stations_coords$start_lat, na.rm = TRUE), zoom = 12)
```