---
title: "Data Transformation"
author: "Danielle Herrington"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Transformation

## 1.	Convert started_at and ended_at to Date-Time Objects

The started_at and ended_at columns were converted to Date-Time objects in YYYY-MM-DD HH:MM:SS format using the lubridate::ymd_hms() function. This ensures that the date-time fields are stored in a format that supports time-based calculations and manipulations.

```{r convert to date-time objects, echo=TRUE}
bike_rides$started_at <- lubridate::ymd_hms(bike_rides$started_at) #turns 'started_at' to ymd_hms
bike_rides$ended_at <- lubridate::ymd_hms(bike_rides$ended_at) #turns 'ended_at' to ymd_hms
```

## 2.	View the Updated Data Structure
The started_at and ended_at fields were confirmed to have been successfully converted to POSIXct date-time objects.


```{r view updated data structure, echo=TRUE}
str(bike_rides)
```

## 3.	Create Hour Fields (start_hour and end_hour)
The hour component (0–23) was extracted from the started_at and ended_at columns and stored in new fields (start_hour and end_hour). This enables analysis of ride patterns by time of day, such as identifying peak usage hours.

```{r view updated data structure, echo=TRUE}
bike_rides$start_hour <- lubridate::hour(bike_rides$started_at) #creates field for 'start hour'
bike_rides$end_hour <- lubridate::hour(bike_rides$ended_at) #creates field for 'end hour'
```

## 4.	Create Date-Only Fields (start_date and end_date)
The date (without the time component) was extracted from started_at and ended_at, storing them in new columns. This allows for day-level analysis, such as tracking usage trends over calendar dates.

```{r create date-only fields, echo=TRUE}
bike_rides$start_date <- as.Date(bike_rides$started_at) #creates field for 'start date'
bike_rides$end_date <- as.Date(bike_rides$ended_at) #creates field for 'end date'
```


## 5.	Add a day_of_week Column
A new column (day_of_week) was created to indicate the day of the week (e.g., "Sunday," "Monday") for each ride's start time. This is useful for understanding weekday vs. weekend usage patterns.

```{r create day_of_week column, echo=TRUE}
bike_rides <- bike_rides %>%
  mutate(day_of_week = wday(started_at, label = TRUE, abbr = FALSE))
```

## 6.	Create a ride_length Column
Ride duration in seconds was calculated by subtracting started_at from ended_at and storing it as a numeric column (ride_length). This allows for detailed analysis of ride durations and helps identify anomalies, such as negative values.

```{r create ride_length column, echo=TRUE}
bike_rides$ride_length <- as.numeric(difftime(bike_rides$ended_at, bike_rides$started_at, units = "secs"))  # Difference in seconds
```

## 7.	Verify the New Columns
The first few rows of the data frame were displayed to confirm that the new columns were added correctly.

```{r verify new columns, echo=TRUE}
head(bike_rides)
```

## 8.	Count Negative Ride Lengths
The number of rides with negative durations (where ride_length is less than 0) was counted and the result printed. Negative ride lengths indicate data issues, such as test rides or errors.

```{r count negative ride lengths, echo=TRUE}
negative_ride_lengths_count <- sum(bike_rides$ride_length < 0, na.rm = TRUE)

print(negative_ride_lengths_count) #Displays the result of negative ride lengths. The data frame includes 258 entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
```

## 9.	Remove Rows with Negative Ride Lengths
Rows where ride_length is less than 0 were filtered out, creating a new cleaned dataset (bike_rides_cleaned). This ensures that the data is accurate and representative of real rides.

```{r remove rows with negative ride lengths, echo=TRUE}
bike_rides_cleaned <- bike_rides %>%
  filter(ride_length >= 0)
```

## 10.	Verify the Cleaned Data
The first few rows of the cleaned data were displayed to verify that negative ride lengths had been removed.

```{r verify cleaned data, echo=TRUE}
head(bike_rides_cleaned)
```

This process prepares the data for meaningful analysis, ensuring that all necessary variables are present and properly formatted to derive insights effectively.
